---
title: "Sample-Size Effects for Projection Pursuit Indexes under Gaussian Noise"
author: "Tina Rashid Jafari"
vignette: >
  %\VignetteIndexEntry{Sample-Size Effects for Projection Pursuit Indexes under Gaussian Noise}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message=FALSE, 
  warning=FALSE
)
```

## Overview

This demonstrates how to use the `ppi_samplesize_effect()` function to study how projection pursuit index (PPI) values behave under increasing sample sizes in purely noisy (Gaussian) data. 
Understanding how an index behaves under null conditions is critical for:

- Calibrating thresholds for statistical significance
- Identifying sensitivity to sample size
- Comparing robustness across different index functions


## What the Function Does

The function `ppi_samplesize_effect()` runs repeated simulations of standard normal noise across a range of sample sizes. For each size, it computes a specified index function over the samples and records the 95th percentile of index values observed. This helps estimate the distributional behavior of the index under pure noise, depending on sample size. These curves are useful as a null reference: above-curve values may indicate non-random structure.




```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(spinebil)
library(cassowaryr)
```

## Function usage

```r
ppi_samplesize_effect(index_fun, n_sim = 100)
```

* `index_fun`: a function that accepts the data for one simulation and returns a single numeric index value. 
* `n_sim`: the number of simulations per sample size (default 100).

> Internally, the function evaluates a grid of sample sizes from 30 up to 2000
> (denser at small sizes), simulates `n_sim` draws for each, and returns the
> 95th percentile of the resulting index values.

## Example Usage

Let’s evaluate the `stringy` index from the `cassowaryr` package to see how its values change under noise as the sample size increases.

```{r}
result <- ppi_samplesize_effect(scagIndex("stringy"), n_sim = 4)
head(result)
tail(result)
```

## Plot the 95th-percentile

```{r fig.height=4.5, fig.width=7.5}
ggplot(result, aes(SampleSize, Percentile95)) +
  geom_point(alpha = 0.7, size = 1) +
  geom_smooth(se = FALSE, method = "loess", formula = y ~ x, span = 0.1) +
  labs(
    x = "Sample size (n)",
    y = "95th percentile of index"
  ) 
```


## Return value

`ppi_samplesize_effect()` returns a tibble:
  
  * `SampleSize`: the sample size for that block of simulations
* `Percentile95`: the empirical 95th percentile of the index across the `n_sim`
simulations at that size.

