---
title: "Generating Basis Vectors and 2D Basis"
author: "Tina Rashid Jafari"
vignette: >
  %\VignetteIndexEntry{Generating Basis Vectors and 2D Basis}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message=FALSE, 
  warning=FALSE
)
```

## Overview

This vignette introduces two utilities:
  
  * `basisVector(i, n)` — the i-th standard basis vector in $\mathbb{R}^n$ (all zeros except a 1 at position `i`).
* `basisMatrix(i, j, n)` — a 2D basis matrix in $\mathbb{R}^n$ whose columns are the `i`-th and `j`-th standard basis vectors.

```{r}
library(spinebil)
library(ggplot2)
```

## What the Function Does

### `basisVector(i, n)`

Returns a length-`n` numeric vector with value `1` at index `i` and `0` elsewhere.

### `basisMatrix(i, j, n)`

Returns an `n × 2` matrix whose first column is `basisVector(i, n)` and second column is `basisVector(j, n)`.


## Function usage

```{r}
# Single direction
e3 <- basisVector(i = 3, n = 5)
e3
```

```{r}
# 2D basis using directions 1 and 3:
B13 <- basisMatrix(i = 1, j = 3, n = 4)
B13
```


## Return value

* **`basisVector(i, n)`** returns numeric vector of length `n` with a single 1 at position `i`.
* **`basisMatrix(i, j, n)`** returns numeric matrix of size `n × 2`.

